Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2011-05-08 release_version/main.tex
Moving release materials into one directory

\begin{figure}
\centering \begin{tabular}{cc}
\includegraphics[height=1in]{figures_cvpr/promo/case1/detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case1/sci_with_axis_face_case1.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/alignment_and_detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case2/sci_with_axis_face_case2.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/case3/alignment.png} &
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case3/sci_with_axis_face_case3.png}
\end{tabular} \caption{{\bf Effects of registration and illumination on
Recognition}. In this example we identify the girl among 20 subjects, by
computing the sparse representation of her input face with respect to the
entire training set. The absolute sum of the coefficients associated with each
subject is plotted on the right. We also show the faces reconstructed with each
subject's training images weighted by the associated sparse coefficients. The
red line (cross) corresponds to her true identity, subject 12. {\bf Top:} The
input face is from Viola and Jones' face detector (the black box) and all 38
illuminations specified in Section \ref{sec:illumination} are used in the
training.  {\bf Middle:} The input face is well-aligned (the white box) with
the training by our algorithm specified in Section \ref{sec:registration} but
only 24 frontal illuminations are used in the training for recognition (see
Section \ref{sec:illumination}). {\bf Bottom:} The input face is well aligned and
a sufficient set (all 38) of
illuminations are used in the training. Both are necessary for correct recognition
using SRC.}\label{fig:promo}
\vspace{-6.9mm}
\end{figure}

@@

\begin{figure}
\centering
{\small
\begin{tabular}{cccc}
\includegraphics[height=.8in]{figures_cvpr/L1_cropped} &
\includegraphics[height=.8in]{figures_cvpr/y_warp_L1} &
\includegraphics[height=.8in]{figures_cvpr/y_hat_L1} &
\includegraphics[height=.8in]{figures_cvpr/e_L1} \\
\includegraphics[height=.8in]{figures_cvpr/L2_cropped} &
\includegraphics[height=.8in]{figures_cvpr/y_warp_L2} &
\includegraphics[height=.8in]{figures_cvpr/y_hat_L2} &
\includegraphics[height=.8in]{figures_cvpr/e_L2} \\
(a) & (b) & (c) & (d)
\end{tabular}
\vspace{0mm}}
\caption{{\bf Comparing alignment of a subject wearing sunglasses by
$\ell^1$ and $\ell^2$ minimization.}
{\bf Top:} alignment result of minimizing $\|\e\|_1$; {\bf Bottom:}
result of minimizing $\|\e\|_2$. (a) {\em Green (dotted):} Initial face boundary
given by the face detector, {\em Red (solid):} Alignment result shown on the same
face; (b) warped testing image using the estimated transformation $\y_0$;
(c) reconstructed face $A_i\x$ using the training; (d) image of error $\e$. }\label{fig:L1-L2-align}
\vspace{-8.0mm}
\end{figure}

@@

\subsection{System Implementation}
The runtime of Algorithm~\ref{alg:deformable-src} is dominated
by the time spent solving two qualitatively similar $\ell_1$ minimization problems.
We have developed custom solvers for this purpose based on 
\emph{Augmented Lagrange Multiplier} (ALM) algorithm.  
We have selected this algorithm because it strikes the best
balance between speed, accuracy, and scalability for our problem out of
many algorithms that we have tested. We refer the reader to our
supplementary materials for a more in-depth discussion of our
solvers.  
For a more detailed discussion of competing
approaches, we refer the interested reader to \cite{YangA2010-pp} 
On a Mac Pro with
Dual-Core 2.66GHz Xeon processors and 4GB memory,
running on our database containing images size $80\times 60$
pixels from 109 subjects under 38 illuminations,
our C implementation of Algorithm~\ref{alg:deformable-src} takes
about 0.60 seconds per subject for alignment and about 2.0
seconds for global recognition. Compared to the highly
customized interior point method used in the conference version
of this paper \cite{Wagner2009-CVPR}, this new algorithm is
only slightly faster for per subject alignment. However, it is
much simpler to implement and it achieves a
\emph{speedup of more than a factor of 10} for global
recognition!

@@

We have gathered the statistics of the Viola and Jones'
 face detector on the Multi-PIE dataset. For 4,600 frontal
 images of 230 subjects under 20 different illuminations,
 using manual registration as the ground truth, the average
 misalignment error of the detected faces is about 6 pixels
 and the average variation in scale is 8\%. This falls
 safely inside the region of attraction for our alignment
 algorithm.
\begin{figure*}
\centering
{\small
\begin{tabular}{ccc}
\includegraphics[height=1.2in]{figures_pami/x_y_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_theta_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_s_roa.png}\\
(a)&(b)&(c)
\end{tabular}
\begin{tabular}{cccc}
\includegraphics[height=1in]{figures_pami/x_tr.png} &
\includegraphics[height=1in]{figures_pami/y_tr.png} &
\includegraphics[height=1in]{figures_pami/theta.png} &
\includegraphics[height=1in]{figures_pami/scale.png}\\
(d)&(e)&(f)& (g)
\end{tabular}
\vspace{0mm}
}
\caption{{\bf Region of attraction.} Fraction of subjects for which the algorithm successfully aligns a synthetically perturbed test image.  The amount of translation is expressed as a fraction of the distance between the outer eye corners, and the amount of in-plane rotation in degrees. {\bf Top row:} (a) Simultaneous translation in $x$ and $y$ directions. (b) Simultaneous translation in $y$ direction and in-plane rotation. (c) Simultaneous translation in $y$ direction and scale variation. {\bf Bottom row:} (d) Translation in $x$ direction only. (e) Translation in $y$ direction only. (f) In-plane rotation only. (g) Scale variation only.}
\label{fig:attraction}
\vspace{0mm}
\end{figure*}

\noindent{2) \em Multiscale Implementation.}
Performing alignment in a multiscale fashion has two benefits: first, it provides a larger region of attraction, and second, it reduces overall computational cost. Here, we further investigate the convergence behavior of the algorithm as a function of the standard deviation $\sigma$ of the Gaussian smoothing filter and the number of scales considered.
We use the same 7 illuminations in
Session 1 as training, and all 20 illuminations in the same
session as testing. We introduce artificial deformation in
both $x$ and $y$ directions up to 16 pixels in the
$80\times 60$ frame, with a step size of 4 pixels, i.e.,
$(\Delta x, \Delta y) \in \{-16,-12,\ldots,12,16\} \times
\{-16,-12,\ldots,12,16\}$. We consider an alignment
successful if the estimated coordinates of the eye-corners
are within 1 pixel from the ground truth in the original
image.  In Figure \ref{fig:multiscale}, we report the
alignment success rate, averaged over the artificially
perturbed initial deformations, as a function of the
standard deviation of the Gaussian kernel $\sigma$, for
three choices of the number of scales. As one can see,
using multiscale indeed improves the performance, and when
3 scales are used, a smaller convolution kernel can achieve
a similar performance compared to a much larger kernel when
only 2 scales are used.
\begin{figure}
\centering
\includegraphics[width=3in]{figures_pami/multiscale.png}
\vspace{-2mm}
\caption{{\bf Multiscale alignment.} This figure shows the average success rate of alignment over all possible perturbations. A smaller blur kernel can be applied to achieve certain level of performance when more scales are used.}
\label{fig:multiscale}
\vspace{-4mm}
\end{figure}

@@

}\label{fig:pose-alignment} \vspace{-4mm}

@@

We have designed a system that can acquire frontal images of a subject while
simultaneously illuminating the subject from all directions above horizontal. A sketch of the
system is shown in Figure \ref{fig:system}: The illumination
system consists of four projectors that display various bright
patterns onto the three white walls in the corner of a dark
room.  The light reflects off of the walls and illuminates the
user's head indirectly.  After taking the frontal illuminations
we rotate the chair by 180 degrees and take pictures from the
opposite direction.  Having two cameras speeds the process
since only the chair needs to be moved in between frontal and
rear illuminations. Our projector-based system has several
advantages over flash-based illumination systems for face recognition:
\begin{itemize}
\item The illuminations can be modified in software, rather than hardware.
\item It is easy to capture many different illuminations quickly.
\item Good coverage and distant illumination can be achieved simultaneously.
\item There is no need to mount anything on the walls or construct a large dome.
\item The system can be assembled from off-the-shelf hardware.
\end{itemize}
\begin{figure}
\centering
\includegraphics[height=2in]{figures_cvpr/camera_rig.pdf}
\caption{{\bf Training acquisition system:} Four projectors and two cameras controlled by one computer.}
\vspace{-2em}
\label{fig:system}
\end{figure}
With our projector system, our choice of illuminations is
constrained only by the need to achieve a good
SNR\footnote{Since illuminations with more pixels illuminated
will have a better SNR (provided they don't saturate), there is
an engineering tradeoff between the SNR and the number of
training images.}, avoid saturation, and achieve
a reasonably short acquisition time.  Two simplifying
assumptions that we make are that every pixel is either turned
fully on or off in every illumination, and that the illuminated
regions do not overlap.

@@

\begin{figure}
\centering
\begin{tabular}{@{}c@{}c@{}}
\includegraphics[height=1.2in]{figures_cvpr/illum_results/coverage_sunset.pdf} &
\includegraphics[height=1.2in]{figures_cvpr/illum_results/granularity_sunset.pdf} \\
(a) Coverage & (b) Granularity
\end{tabular}\vspace{0mm}
\caption{{\bf Study of sufficient illuminations.} The average $\ell^1$ registration residual versus different illumination training sets. }
\label{fig:illumination-sufficiency}
\vspace{-4.5em}
\end{figure}

@@

\newcommand{\tempwidth}[0]{0.58in}
\begin{figure}
\centering
{\small
\begin{tabular}{@{}c@{}c@{}c@{}c@{}c@{}c@{}}
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \\
\hspace{-2mm}(a) & (b) & (c) & (d) & (e) & (f) \vspace{0mm}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-4mm}
\end{figure}

\subsection{Improving the Sampling Window}
Our algorithm's errors are mostly caused by a few subjects who
significantly change their appearances between sessions (such
as hair, facial hair, and eyeglasses). Some representative
examples are shown in Figure \ref{fig:failed-examples}. For those subjects, alignment and recognition fail on
almost all test illuminations.

Meanwhile, this observation also suggests that we might be able
to improve the performance of our method by carefully choosing
a face region which is less affected by the above factors for
recognition. In particular, since the forehead region is likely
to be affected by the change of hair style, we try replacing
the previous $80 \times 60$ canonical frame with a new
window that better excludes the forehead. We adjust the
resolution of the window to keep $m$ approximately constant. In addition,
we cut off two lower corners of the $80 \times 60$ canonical frame, motivated by
the observation that in many cases the corners
actually contain background. An example of the new window
is shown in Figure~\ref{fig:new-mask}.

\begin{figure}
\vspace{-1em}
\centering
{\small
\begin{tabular}{@{}cc@{}}
\includegraphics[trim=1.9in .7in 1.9in .5in, clip, height=1.0in]{figures_pami/example.png} &
\includegraphics[trim=1.9in .7in 1.9in .5in, clip, height=1.0in]{figures_pami/example_new.png} \vspace{0mm} \\
Default window. & Proposed window. \vspace{0mm}
\end{tabular}
}
\caption{{\bf Choosing different sampling windows.}}
\label{fig:new-mask}
\vspace{-1em}
\end{figure}

@@

\subsection{Comparison to Existing Work}
We first compare our result to the recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is obtained from manually selected outer
eye corners. Then, a supervised hierarchical sparse coding
model based on local image descriptors is trained, which enjoys
certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle the remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition2} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

@@

% LBP
We also compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is
able to capture fine details of facial appearance and texture.
Due to its robustness to variations in illumination, facial
expression, aging and other changes, LBP has achieved the
state-of-the-art face recognition performance in the scenario
when only one sample per person is used for training
\cite{Tan06facerecognition}.
In this paper, we follow the same
steps as in \cite{Ahonen2006-PAMI} to construct an LBP
descriptor for each training and testing sample. The $80\times
60$ face region is first divided into a regular $10\times 10$
grid of cells, each of size $8\times 6$ pixels. Within each
cell, the histogram of 59 uniform binary patterns is then
computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central
pixel value. Finally, the local histograms are concatenated to
produce the global descriptor vector. As suggested in
\cite{Ahonen2006-PAMI}, the recognition is performed using a
nearest neighbor classifier with Chi square distance as the
distance measure and we report the recognition rates with the
same three types of input as before.  As shown in
Table~\ref{tab:MultiPIE-recognition}, although LBP achieves
competitive recognition rates given manually aligned training
and testing samples, demonstrating its robustness to moderate
misalignment, it still benefits from using the output of our
iterative alignment algorithm as the input.
In addition, like
the other classical algorithms, the performance of LBP degrades
dramatically if it is applied directly to the output of a face
detector. This is notable given that LBP is often applied
without any special alignment in practice. Finally, we
attribute the improvement in performance of LBP over SRC in
this experiment to its robustness to illumination components
that cannot be linearly interpolated by the training set.

@@

\begin{table}
\caption{Performance on single gallery image FERET dataset}
\vspace{-3mm} \centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate \% & fb & fc & dup1 & dup2 \\
\hline
$LBP_d$ & 54.8 & 10.3 & 29.8 & 19.8 \\
\hline
$LBP_m$ & {\bf 96.6} & {\bf 58.8} & {\bf 71.6} & {\bf 61.5} \\
\hline
$LBP_i$ & 94.5 & 42.8 & 46.5 & 21.1 \\
\hline
{Alg. 1} & 95.2 & 28.4 & 46.1 & 20.3 \\
\hline
\end{tabular}
\label{tab:FERET-recognition} }
\vspace{-12mm}
\end{table}

As expected, our algorithm does not perform well except for
`fb', in which the illumination is similar to the training and
the mere variation in facial expression is handled well by the
sparse error model. For the other three test sets, our
algorithm fails because the illumination changes and other
variations seriously violate the assumptions of our method.
This also explains why LBP performs worse with our iterative
alignment algorithm, compared to manual alignment. On the other
hand, while LBP achieves the best recognition rates given
manually aligned training and testing samples, its performance
degrades drastically when the input is obtained directly from
the face detector. It is also worth noting that similar poor
performance of LBP, as well as other descriptors, has been
observed on the Labeled Face in the Wild (LFW) database, where
the training is uncontrolled and limited and the input is
directly obtained from the face detector \cite{Wolf2008-ECCV}.

@@

\begin{figure}
\centering
\includegraphics[width=2.6in]{figures_pami/occ_block.png}
\caption{{\bf Using overlapping blocks to tackle contiguous occlusion.} (a) The test image, occluded by sunglasses. (b) The four overlapping blocks. (c) The sparse representation is calculated after alignment for each block independently. The red lines correspond to his true identity. (d) The true identity is successfully recovered by voting based on the SCI scores.}
\label{fig:occ-block}
\vspace{-3em}
\end{figure}

@@

\input{appendix}

>>>>

\begin{figure}
\centering \begin{tabular}{cc}
\includegraphics[height=1in]{figures_cvpr/promo/case1/detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case1/sci_with_axis_face_case1.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/alignment_and_detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case2/sci_with_axis_face_case2.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/case3/alignment.png} &
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case3/sci_with_axis_face_case3.png}
\end{tabular} \caption{{\bf Effects of registration and illumination on
Recognition}. In this example we identify the girl among 20 subjects, by
computing the sparse representation of her input face with respect to the
entire training set. The absolute sum of the coefficients associated with each
subject is plotted on the right. We also show the faces reconstructed with each
subject's training images weighted by the associated sparse coefficients. The
red line (cross) corresponds to her true identity, subject 12. {\bf Top:} The
input face is from Viola and Jones' face detector (the black box) and all 38
illuminations specified in Section \ref{sec:illumination} are used in the
training.  {\bf Middle:} The input face is well-aligned (the white box) with
the training by our algorithm specified in Section \ref{sec:registration} but
only 24 frontal illuminations are used in the training for recognition (see
Section \ref{sec:illumination}). {\bf Bottom:} The input face is well aligned and
a sufficient set (all 38) of
illuminations are used in the training. Both are necessary for correct recognition
using SRC.}\label{fig:promo}
\vspace{-5mm}
\end{figure}

@@

\begin{figure}
\centering
{\small
\begin{tabular}{cccc}
\includegraphics[height=.8in]{figures_cvpr/L1_cropped} &
\includegraphics[height=.8in]{figures_cvpr/y_warp_L1} &
\includegraphics[height=.8in]{figures_cvpr/y_hat_L1} &
\includegraphics[height=.8in]{figures_cvpr/e_L1} \\
\includegraphics[height=.8in]{figures_cvpr/L2_cropped} &
\includegraphics[height=.8in]{figures_cvpr/y_warp_L2} &
\includegraphics[height=.8in]{figures_cvpr/y_hat_L2} &
\includegraphics[height=.8in]{figures_cvpr/e_L2} \\
(a) & (b) & (c) & (d)
\end{tabular}
\vspace{0mm}}
\caption{{\bf Comparing alignment of a subject wearing sunglasses by
$\ell^1$ and $\ell^2$ minimization.}
{\bf Top:} alignment result of minimizing $\|\e\|_1$; {\bf Bottom:}
result of minimizing $\|\e\|_2$. (a) {\em Green (dotted):} Initial face boundary
given by the face detector, {\em Red (solid):} Alignment result shown on the same
face; (b) warped testing image using the estimated transformation $\y_0$;
(c) reconstructed face $A_i\x$ using the training; (d) image of error $\e$. }\label{fig:L1-L2-align}
\vspace{-5.0mm}
\end{figure}

@@

\subsection{System Implementation}
The runtime of Algorithm~\ref{alg:deformable-src} is dominated
by the time spent solving two qualitatively similar $\ell_1$ minimization problems.
We have developed custom solvers for this purpose based on
\emph{Augmented Lagrange Multiplier} (ALM) algorithm.
We have selected this algorithm because it strikes the best
balance between speed, accuracy, and scalability for our problem out of
many algorithms that we have tested. We refer the reader to our
supplementary materials for a more in-depth discussion of our
solvers.
For a more detailed discussion of competing
approaches, we refer the interested reader to
\cite{YangA2010-pp}.
On a Mac Pro with
Dual-Core 2.66GHz Xeon processors and 4GB memory,
running on our database containing images size $80\times 60$
pixels from 109 subjects under 38 illuminations,
our C implementation of Algorithm~\ref{alg:deformable-src} takes
about 0.60 seconds per subject for alignment and about 2.0
seconds for global recognition. Compared to the highly
customized interior point method used in the conference version
of this paper \cite{Wagner2009-CVPR}, this new algorithm is
only slightly faster for per subject alignment. However, it is
much simpler to implement and it achieves a
\emph{speedup of more than a factor of 10} for global
recognition!

@@

We have gathered the statistics of the Viola and Jones'
 face detector on the Multi-PIE dataset. For 4,600 frontal
 images of 230 subjects under 20 different illuminations,
 using manual registration as the ground truth, the average
 misalignment error of the detected faces is about 6 pixels
 and the average variation in scale is 8\%. This falls
 safely inside the region of attraction for our alignment
 algorithm.
\begin{figure*}
\centering
{\small
\begin{tabular}{ccc}
\includegraphics[height=1.2in]{figures_pami/x_y_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_theta_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_s_roa.png}\\
(a)&(b)&(c)
\end{tabular}
\begin{tabular}{cccc}
\includegraphics[height=1in]{figures_pami/x_tr.png} &
\includegraphics[height=1in]{figures_pami/y_tr.png} &
\includegraphics[height=1in]{figures_pami/theta.png} &
\includegraphics[height=1in]{figures_pami/scale.png}\\
(d)&(e)&(f)& (g)
\end{tabular}
\vspace{0mm} } \caption{{\bf Region of attraction.} Fraction of
subjects for which the algorithm successfully aligns a
synthetically perturbed test image.  The amount of translation
is expressed as a fraction of the distance between the outer
eye corners, and the amount of in-plane rotation in degrees.
{\bf Top row:} (a) Simultaneous translation in $x$ and $y$
directions. (b) Simultaneous translation in $y$ direction and
in-plane rotation. (c) Simultaneous translation in $y$
direction and scale variation. {\bf Bottom row:} (d)
Translation in $x$ direction only. (e) Translation in $y$
direction only. (f) In-plane rotation only. (g) Scale variation
only.} \label{fig:attraction} \vspace{-4mm}
\end{figure*}

\noindent{2) \em Multiscale Implementation.}
Performing alignment in a multiscale fashion has two benefits: first, it provides a larger region of attraction, and second, it reduces overall computational cost. Here, we further investigate the convergence behavior of the algorithm as a function of the standard deviation $\sigma$ of the Gaussian smoothing filter and the number of scales considered.
We use the same 7 illuminations in
Session 1 as training, and all 20 illuminations in the same
session as testing. We introduce artificial deformation in
both $x$ and $y$ directions up to 16 pixels in the
$80\times 60$ frame, with a step size of 4 pixels, i.e.,
$(\Delta x, \Delta y) \in \{-16,-12,\ldots,12,16\} \times
\{-16,-12,\ldots,12,16\}$. We consider an alignment
successful if the estimated coordinates of the eye-corners
are within 1 pixel from the ground truth in the original
image.  In Figure \ref{fig:multiscale}, we report the
alignment success rate, averaged over the artificially
perturbed initial deformations, as a function of the
standard deviation of the Gaussian kernel $\sigma$, for
three choices of the number of scales. As one can see,
using multiscale indeed improves the performance, and when
3 scales are used, a smaller convolution kernel can achieve
a similar performance compared to a much larger kernel when
only 2 scales are used.
\begin{figure}
\centering
\includegraphics[width=3in]{figures_pami/multiscale.png}
\vspace{-2mm}
\caption{{\bf Multiscale alignment.} This figure shows the average success rate of alignment over all possible perturbations. A smaller blur kernel can be applied to achieve certain level of performance when more scales are used.}
\label{fig:multiscale}
\vspace{-2mm}
\end{figure}

@@

}\label{fig:pose-alignment} \vspace{-8mm}

@@

We have designed a system that can acquire frontal images of a subject while
simultaneously illuminating the subject from all directions above horizontal. A sketch of the
system is shown in Figure \ref{fig:system}: The illumination
system consists of four projectors that display various bright
patterns onto the three white walls in the corner of a dark
room.  The light reflects off of the walls and illuminates the
user's head indirectly.  After taking the frontal illuminations
we rotate the chair by 180 degrees and take pictures from the
opposite direction.  Having two cameras speeds the process
since only the chair needs to be moved in between frontal and
rear illuminations. Our projector-based system has several
advantages over flash-based illumination systems for face recognition:
\begin{itemize}
\item The illuminations can be modified in software, rather than hardware.
\item It is easy to capture many different illuminations quickly.
\item Good coverage and distant illumination can be achieved simultaneously.
\item There is no need to mount anything on the walls or construct a large dome.
\item The system can be assembled from off-the-shelf hardware.
\end{itemize}
\begin{figure}
\centering
\includegraphics[height=2in]{figures_cvpr/camera_rig.pdf}
\caption{{\bf Training acquisition system:} Four projectors and two cameras controlled by one computer.}
\vspace{-2em}
\label{fig:system}
\end{figure}
With our projector system, our choice of illuminations is
constrained only by the need to achieve a good
SNR,\footnote{Since illuminations with more pixels illuminated
will have a better SNR (provided they don't saturate), there is
an engineering tradeoff between the SNR and the number of
training images.} avoid saturation, and achieve a reasonably
short acquisition time.  Two simplifying assumptions that we
make are that every pixel is either turned fully on or off in
every illumination, and that the illuminated regions do not
overlap.

@@

\begin{figure}
\centering
\begin{tabular}{@{}c@{}c@{}}
\includegraphics[height=1.2in]{figures_cvpr/illum_results/coverage_sunset.pdf} &
\includegraphics[height=1.2in]{figures_cvpr/illum_results/granularity_sunset.pdf} \\
(a) Coverage & (b) Granularity
\end{tabular}\vspace{0mm}
\caption{{\bf Study of sufficient illuminations.} The average $\ell^1$ registration residual versus different illumination training sets. }
\label{fig:illumination-sufficiency}
\vspace{-3em}
\end{figure}

@@

\newcommand{\tempwidth}[0]{0.58in}
\begin{figure}
\centering
{\small
\begin{tabular}{@{}c@{}c@{}c@{}c@{}c@{}c@{}}
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \\
\hspace{-2mm}(a) & (b) & (c) & (d) & (e) & (f) \vspace{0mm}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-12mm}
\end{figure}
\subsection{Improving the Sampling Window}
Our algorithm's errors are mostly caused by a few subjects who
significantly change their appearances between sessions (such
as hair, facial hair, and eyeglasses). Some representative
examples are shown in Figure \ref{fig:failed-examples}. For those subjects, alignment and recognition fail on
almost all test illuminations.
\begin{figure}[b]
\vspace{-1em}
\centering
{\small
\begin{tabular}{@{}cc@{}}
\includegraphics[trim=1.9in .7in 1.9in .5in, clip, height=1.0in]{figures_pami/example.png} &
\includegraphics[trim=1.9in .7in 1.9in .5in, clip, height=1.0in]{figures_pami/example_new.png} \vspace{0mm} \\
Default window. & Proposed window. \vspace{0mm}
\end{tabular}
}
\caption{{\bf Choosing different sampling windows.}}
\label{fig:new-mask}
\vspace{-1em}
\end{figure}
Meanwhile, this observation also suggests that we might be able
to improve the performance of our method by carefully choosing
a face region which is less affected by the above factors for
recognition. In particular, since the forehead region is likely
to be affected by the change of hair style, we try replacing
the previous $80 \times 60$ canonical frame with a new
window that better excludes the forehead. We adjust the
resolution of the window to keep $m$ approximately constant. In addition,
we cut off two lower corners of the $80 \times 60$ canonical frame, motivated by
the observation that in many cases the corners
actually contain background. An example of the new window
is shown in Figure~\ref{fig:new-mask}.

@@

\subsection{Comparison to Existing Work}

We first compare our result to the recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is obtained from manually selected outer
eye corners. Then, a supervised hierarchical sparse coding
model based on local image descriptors is trained, which enjoys
certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle the remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition2} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

@@

% LBP
We also compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is
able to capture fine details of facial appearance and texture.
Due to its robustness to variations in illumination, facial
expression, aging and other changes, LBP has achieved the
state-of-the-art face recognition performance in the scenario
when only one sample per person is used for training
\cite{Tan06facerecognition}. In this paper, we follow the same
steps as in \cite{Ahonen2006-PAMI} to construct an LBP
descriptor for each training and testing sample. The $80\times
60$ face region is first divided into a regular $10\times 10$
grid of cells, each of size $8\times 6$ pixels. Within each
cell, the histogram of 59 uniform binary patterns is then
computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central
pixel value. Finally, the local histograms are concatenated to
produce the global descriptor vector. As suggested in
\cite{Ahonen2006-PAMI}, the recognition is performed using a
nearest neighbor classifier with Chi square distance as the
distance measure and we report the recognition rates with the
same three types of input as before.

As shown in Table~\ref{tab:MultiPIE-recognition}, although LBP
achieves competitive recognition rates given manually aligned
training and testing samples, demonstrating its robustness to
moderate misalignment, it still benefits from using the output
of our iterative alignment algorithm as the input. In addition,
like the other classical algorithms, the performance of LBP
degrades dramatically if it is applied directly to the output
of a face detector. This is notable given that LBP is often
applied without any special alignment in practice. Finally, we
attribute the improvement in performance of LBP over SRC in
this experiment to its robustness to illumination components
that cannot be linearly interpolated by the training set.

@@

As expected, our algorithm does not perform well except for
`fb', in which the illumination is similar to the training and
the mere variation in facial expression is handled well by the
sparse error model. For the other three test sets, our
algorithm fails because the illumination changes and other
variations seriously violate the assumptions of our method.
This also explains why LBP performs worse with our iterative
alignment algorithm, compared to manual alignment. On the other
hand, while LBP achieves the best recognition rates given
manually aligned training and testing samples, its performance
degrades drastically when the input is obtained directly from
the face detector. It is also worth noting that similar poor
performance of LBP, as well as other descriptors, has been
observed on the Labeled Face in the Wild (LFW) database, where
the training is uncontrolled and limited and the input is
directly obtained from the face detector \cite{Wolf2008-ECCV}.
\begin{table}
\caption{Performance on single gallery image FERET dataset}
\vspace{-3mm} \centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate \% & fb & fc & dup1 & dup2 \\
\hline
$LBP_d$ & 54.8 & 10.3 & 29.8 & 19.8 \\
\hline
$LBP_m$ & {\bf 96.6} & {\bf 58.8} & {\bf 71.6} & {\bf 61.5} \\
\hline
$LBP_i$ & 94.5 & 42.8 & 46.5 & 21.1 \\
\hline
{Alg. 1} & 95.2 & 28.4 & 46.1 & 20.3 \\
\hline
\end{tabular}
\label{tab:FERET-recognition} }
\vspace{-12mm}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=2.6in]{figures_pami/occ_block.png}
\vspace{-2mm}
\caption{{\bf Using overlapping blocks to tackle contiguous occlusion.} (a) The test image, occluded by sunglasses. (b) The four overlapping blocks. (c) The sparse representation is calculated after alignment for each block independently. The red lines correspond to his true identity. (d) The true identity is successfully recovered by voting based on the SCI scores.}
\label{fig:occ-block}
\vspace{-5mm}
\end{figure}

@@

% \input{appendix_main}
